---
title: "Group_5_Analysis"
format: html
editor: visual
---

## Introduction

## Setup

```{r}
library(tidyverse)
library(janitor)
library(skimr)
library(GGally)
library(MASS)
library(AER)
library(car)
library(broom)
library(performance)
library(pscl)
library(knitr)
library(kableExtra)
library(ggplot2)
library(patchwork)
```

## Data description

```{r}
df <- read_csv("D:/dataset05.csv")

glimpse(df)
summary(df)
```

## Data cleaning and preprocessing

### Initial checks

```{r}
dim(df)
names(df)
colSums(is.na(df))
```

### Check unique values for categorical variables

```{r}
df %>% count(Region, sort = TRUE)
df %>% count(Household.Head.Sex, sort = TRUE)
df %>% count(Type.of.Household, sort = TRUE)
df %>% count(Electricity, sort = TRUE)
```

### Convert variable types

```{r}
df_clean <- df %>%
  mutate(
    Region = as.factor(Region),
    Household.Head.Sex = as.factor(Household.Head.Sex),
    Type.of.Household = as.factor(Type.of.Household),
    Electricity = factor(Electricity, levels = c(0, 1), labels = c("No", "Yes"))
  )
```

### Create transformed variables

```{r}
df_clean <- df_clean %>%
  mutate(
    log_income = log(Total.Household.Income),
    log_food_exp = log(Total.Food.Expenditure),
    log_floor_area = log(House.Floor.Area),
    head_age_sq = Household.Head.Age^2
  )
```

## Exploratory data analysis

### Response variable distribution

```{r}
df_clean %>%
  summarise(
    mean_family_size = mean(Total.Number.of.Family.members),
    var_family_size = var(Total.Number.of.Family.members),
    min_family_size = min(Total.Number.of.Family.members),
    max_family_size = max(Total.Number.of.Family.members)
  ) %>%
  kable()
```

```{r}
ggplot(df_clean, aes(x = Total.Number.of.Family.members)) +
  geom_bar() +
  labs(
    title = "Distribution of household size",
    x = "Number of family members",
    y = "Count"
  )
```

### Numeric variable distributions

```{r}
p_income_raw <- ggplot(df_clean, aes(x = Total.Household.Income)) +
  geom_histogram(bins = 30) +
  labs(title = "Original household income", x = "Total household income", y = "Frequency")

p_income_log <- ggplot(df_clean, aes(x = log_income)) +
  geom_histogram(bins = 30) +
  labs(title = "Log household income", x = "Log household income", y = "Frequency")

p_food_raw <- ggplot(df_clean, aes(x = Total.Food.Expenditure)) +
  geom_histogram(bins = 30) +
  labs(title = "Original food expenditure", x = "Total food expenditure", y = "Frequency")

p_food_log <- ggplot(df_clean, aes(x = log_food_exp)) +
  geom_histogram(bins = 30) +
  labs(title = "Log food expenditure", x = "Log food expenditure", y = "Frequency")

p_area_raw <- ggplot(df_clean, aes(x = House.Floor.Area)) +
  geom_histogram(bins = 30) +
  labs(title = "Original floor area", x = "House floor area", y = "Frequency")

p_area_log <- ggplot(df_clean, aes(x = log_floor_area)) +
  geom_histogram(bins = 30) +
  labs(title = "Log floor area", x = "Log floor area", y = "Frequency")

(p_income_raw | p_income_log) /
(p_food_raw   | p_food_log) /
(p_area_raw   | p_area_log)
```

### Household size by categorical predictors

```{r}
ggplot(df_clean, aes(x = Household.Head.Sex, y = Total.Number.of.Family.members)) +
  geom_boxplot() +
  labs(title = "Household size by sex of household head")
```

```{r}
ggplot(df_clean, aes(x = Type.of.Household, y = Total.Number.of.Family.members)) +
  geom_boxplot() +
  labs(title = "Household size by type of household") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
ggplot(df_clean, aes(x = Electricity, y = Total.Number.of.Family.members)) +
  geom_boxplot() +
  labs(title = "Household size by electricity access")
```

### Household size against numeric predictors

```{r}
GGally::ggpairs(
  dplyr::select(
    df_clean,
    Total.Number.of.Family.members,
    log_income,
    log_food_exp,
    Household.Head.Age,
    log_floor_area,
    House.Age,
    Number.of.bedrooms
  ),
  upper = list(continuous = "cor"),
  lower = list(continuous = "points")
)
```

## GLM modelling strategy

### Poisson regression

```{r}

```

### Check for overdispersion

```{r}

```

### Quasi-Poisson

```{r}

```

### Negative Binomial

```{r}

```

## Model comparison

```{r}

```

## Diagnostics

```{r}

```

## Conclusion
